---
title: "class_exercises"
author: "Chanuwas (New) Aswamenakul"
date: "1/24/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

# Date: 01/24/2022

## Exercises

```{r}
# 1. Make a vector of "ten" of all the multiples of 10 up to 200
a <- seq(from = 10, to = 200, by = 10)
a

# 2. Find the indices of the numbers divisible by 3
which(a %% 3 == 0)

# 3. Write a function that takes a vector of numbers as an argument and returns that vectors mean and SD
bstats_func <- function(nums) {
  return(list(mean = mean(nums), sd = sd(nums)))
}

# 4. Use your function to get the mean and SD of tens
bstats_func(a)
```

# Date: 01/26/2022

```{r}
rating.df <- read_csv(file.path("data", "sizeRatingsbySub.csv"))
rating.df %>% glimpse()
```

```{r}
rating.df %>% 
  mutate(word.length = if_else(str_length(Word) > 4, "long", "short"))

rating.df %>% 
  group_by(Subject) %>% 
  mutate(avg.rating.by.subj = mean(Rating))
```

## Exercises:

```{r}
# 1. create a new version of the dataset which contains only data for responses to the words: squid, gherkin, mustard
select.responses <- rating.df %>% 
  filter(Word %in% c("squid", "gherkin", "mustard"))

select.responses
```

```{r}
# 2. Add column to the dataframe that codes whether the word ends with "n"
rating.df <- rating.df %>% 
  mutate(endwithn = if_else(str_ends(Word, "n"), TRUE, FALSE))

# 3. Get the mean ratings and counts for words separated by whether they end with "n" or not
rating.df %>% 
  group_by(endwithn) %>% 
  summarize(avg.rating = mean(Rating), count = n())
```

```{r}
fam.df <- read_csv(file.path("data", "meanFamiliarityByItem.csv"))
fam.df %>% glimpse()

joined.df <- rating.df %>% 
  left_join(fam.df, by = "Word")

joined.df %>% glimpse()
```

```{r}
# pivoting example
joined.df %>% 
  group_by(Subject, Class) %>% 
  summarize(avg.rating = mean(Rating)) %>% 
  pivot_wider(names_from = Class, values_from = avg.rating)
```

# Date: 01/31/2022

```{r}
# load diamonds dataset
df.diamonds <- diamonds
df.diamonds %>% glimpse()
```

```{r}
# summarize data
df.diamonds.sum <- df.diamonds %>% 
  group_by(color) %>% 
  summarize(price.mean = mean(price))

# plot stat summary
df.diamonds.sum %>% 
  ggplot(aes(x = color, y = price.mean)) +
  geom_col()

# plot stat summary without group_by and summarize
df.diamonds %>% 
  ggplot(aes(x = color, y = price)) +
  stat_summary(fun = "mean", geom = "col")

# changing theme
theme_set(theme_classic() +
            theme(text = element_text(size = 20)))
df.diamonds %>% 
  ggplot(aes(x = color, y = price)) +
  stat_summary(fun = "mean", geom = "col")

# scatter plot
df.diamonds %>% 
  ggplot(aes(x = carat, y = price, color = color)) +
  geom_point()

# change back to default theme
theme_set(theme_grey())
```

## Exercise

```{r}
# make scatter plot to show the relationship between x = depth and y = table
df.diamonds %>% 
  ggplot(aes(x = depth, y = table)) +
  geom_point()
```

```{r}
# fun.args = argument for fun.data (B = bootstrap iteration)
df.diamonds %>% 
  ggplot(aes(x = clarity, y = price)) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange",
               fun.args = list(conf.int = .999, B = 2000))
```
## Exercise: errorbar

Plot
y = avg price
x = clarity
light blue bars, black outline, add error bars

```{r}
df.diamonds %>% 
  ggplot(aes(x = clarity, y = price)) +
  stat_summary(fun = "mean", geom = "col", color = "black", fill = "lightblue") +
  stat_summary(fun.data = "mean_cl_boot", geom = "errorbar", width = .2) # geom = "linerange" works too
```

```{r}
# group data in plot
df.diamonds %>% 
  ggplot(aes(x = color,
             y = price,
             group = cut,
             color = cut)) +
  stat_summary(fun = "mean", geom = "line", size = 1.25)
```

## Exercise
add error bar to the above plot

```{r}
df.diamonds %>% 
  ggplot(aes(x = color,
             y = price,
             group = cut,
             color = cut)) +
  stat_summary(fun = "mean", geom = "line", size = 1.25) +
  stat_summary(fun.data = "mean_cl_boot", geom = "errorbar", width = .2)
```

```{r}
# bar plot with group and errorbars
df.diamonds %>% 
  ggplot(aes(x = color, y = price, group = cut, fill = cut)) +
  stat_summary(fun = "mean", geom = "col", position = position_dodge()) +
  stat_summary(fun.data = "mean_cl_boot", geom = "errorbar",
               position = position_dodge(.9), width = .2)

# distribution of diamond's width
df.diamonds %>%
  ggplot(aes(x = y)) +
  geom_histogram(bins = 200) +
  coord_cartesian(xlim = c(3, 10)) # zooming in

# adding color to different diamond color
df.diamonds %>% 
  ggplot(aes(x = y, fill = color)) +
  geom_density(bw = .2, alpha = .3) +
  coord_cartesian(xlim = c(3, 10))

# adding facets
df.diamonds %>% 
  ggplot(aes(x = y, fill = color)) +
  geom_density(bw = .2) +
  facet_grid(color~.) +
  coord_cartesian(xlim = c(3, 10))
```

## Exercise: facet wrap

scatterplot for the relationship between carat and price for each color using facetting

```{r}
df.diamonds %>% 
  ggplot(aes(x = carat, y = price, color = color)) +
  geom_point() +
  facet_wrap(color~.)
```
# Date: 02/02/2022

```{r}
df.diamonds %>% 
  ggplot(aes(x = carat, y = price)) + # global aesthetic
  geom_point(aes(color = color)) + # local aesthetic
  geom_smooth(method = "lm", se = FALSE, color = "black")

# jittering points
df.diamonds %>% 
  ggplot(aes(x = color, y = price)) +
  geom_point(alpha = 0.2,
             position = position_jitter(width = 0.1, height = 0),
             size = 2)

df.diamonds[1:150,] %>% 
  ggplot(aes(x = color, y = price)) +
  geom_point(alpha = 0.2,
             position = position_jitter(width = 0.1, height = 0),
             size = 2) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange", fill = "lightblue", shape = 21)
```

## Exercise

y = price, x = cut, individual data points, violin, mean

```{r}
g1 <- df.diamonds %>% 
  ggplot(aes(x = cut, y = price)) +
  geom_point(alpha = .3,
             position = position_jitter(width = .2, height = 0)) +
  geom_violin(alpha = 0.3) +
  stat_summary(fun = "mean", geom = "point",
               color = "red")
g1
```

```{r}
library(ggridges)

g2 <- df.diamonds %>% 
  ggplot(aes(x = price, y = color)) +
  ggridges::geom_density_ridges(scale = 1.5)
g2

# Color brewer helps pick color palette
g3 <- df.diamonds %>% 
  ggplot(aes(x = color, y = clarity, z = carat)) +
  stat_summary_2d(fun = "mean", geom = "tile", color = "black") +
  scale_fill_gradient(low = "lightyellow", high = "red")
g3

ggsave(filename = "plot.pdf", plot = g3, width = 8, height = 6)
```

```{r}
library(patchwork)

# combining plot
g1 + (g2 + g3) +
  plot_layout(ncol = 1) &
  theme_classic() &
  theme(plot.tag = element_text(face = "bold", size = 20))
```

## Distributions

dnorm() = density (not the same as probability for continuous variables)
pnorm() = cumulative distribution
qnorm() = quantile function
rnorm() = generate random values

```{r}
help("Distributions")
```

```{r}
# x = number of positive outcome
# size = total number of trials
dbinom(x = 7, size = 10, prob = .7)

x <- 0:10
p <- dbinom(x, size = 10, prob = .7)

plot(x, p)

# varying parameters
theta <- seq(from = 0, to = 1, by = 0.01)
lh <- dbinom(x = 3, size = 10, prob = theta)

plot(theta, lh)
```

## Exercise: bernoulli

prob = 0.6,
size = 1

```{r}
x <- 0:1
d <- dbinom(x = x, size = 1, prob = 0.6)

plot(x, d)
```

```{r pqrbinom}
x <- 0:10
p <- pbinom(q = x, size = 10, prob = .7)

plot(x, p)

q <- qbinom(p = 0.35, size = 10, prob = .7)
q

x <- rbinom(n = 1000, size = 10, prob = .7)
plot(x)
hist(x)
```

```{r norm}
# normal distribution
dnorm(3.4, m = 0, sd = 1)

x <- seq(from = -5, to = 5, by = 0.1)
d <- dnorm(x, m = 0, sd = 1)

plot(x, d, type = "line")

x <- seq(from = -5, to = 5, by = 0.1)
cp <- pnorm(x, m = 0, sd = 1)

plot(x, cp, type = "line")

pnorm(0)
```

## Exercise: norm

find the probability of an x coming from within 1 SD of the mean, 2 SD, 3 SD

```{r}
# mean = 0, sd = 1
pnorm(1) - pnorm(-1)
pnorm(2) - pnorm(-2)
pnorm(3) - pnorm(-3)
```

# Date: 02/07/2022

```{r}
library(mvtnorm)

x <- seq(from = -5, to = 5, by = .1)
d <- dnorm(x, m = 0, sd = 1)

qplot(x, d) + geom_line() + 
  geom_vline(aes(xintercept = 2), color = "red") +
  geom_vline(aes(xintercept = -2), color = "red")
```

```{r}
# p(x < 2)
pnorm(2, m = 0, sd = 1)

# p(-2 < x < 2)
pnorm(2, m = 0, sd = 1) - pnorm(-2, m = 0, sd = 1)

# quantile function
qnorm(0.977)

# random number generation
rand_nums <- rnorm(2000, 50, 1)
hist(rand_nums)

mean(rand_nums)
sd(rand_nums)
```

## Exercise: t-distribution

- density for a t-distribution with df = 9
- find critical value corresponding to the 97.5th percentile

```{r}
x <- seq(-5, 5, by = .1)
d <- dt(x, df = 9) # density function

qplot(x, d) +
  geom_line()

# critical value at 97.5th percentile
qt(0.975, df = 9)
```

```{r}
# critical value shrinks as df increases (larger sample size)
n <- 10:20
crit.t <- qt(0.975, df = n)

plot(n, crit.t)
```

## Multivariate normal distribution

```{r}
# covariance matrix
sigma <- matrix(c(4, 3, 3, 5), ncol = 2)
sigma

# multivariate normal
x <- mvtnorm::rmvnorm(n = 1000, sigma = sigma)
plot(x)

cor(x)
rho <- cov2cor(sigma) # convert covariance to correlation
rho
sigma[2]/sqrt(sigma[1]*sigma[4])

# using ggplot
as.data.frame(x) %>% 
  ggplot(aes(x = V1, y = V2)) +
  geom_point() +
  geom_density_2d_filled()

x <- mvtnorm::rmvnorm(n = 1000, mean = c(1,2), sigma = rho)
as.data.frame(x) %>% 
  ggplot(aes(x = V1, y = V2)) +
  geom_point() +
  geom_density_2d()
```

## Exercise: multivariate t-dist

- sample 500 points from bivariate t-distribution (df = 9)
- make a 2d density plot

```{r}
sigma <- matrix(c(1, 0, 0, 1), ncol = 2)
x <- mvtnorm::rmvt(500, sigma = sigma, df = 9)

as.data.frame(x) %>% 
  ggplot(aes(x = V1, y = V2)) +
  geom_point() +
  geom_density_2d()
```

## t-test

```{r}
d <- tibble(grp1 = rnorm(20, m = 2, sd = 5),
            grp2 = rnorm(20, m = 5, sd = 5))

d %>% 
  pivot_longer(cols = everything(), names_to = "grp", values_to = "memory") %>% 
  ggplot(aes(x = memory, fill = grp)) +
  geom_histogram(position = position_dodge())

t.test(d$grp1, d$grp2, var.equal = T)

t_value <- (mean(d$grp1) - mean(d$grp2))/(sqrt(var(d$grp1)/length(d$grp1) + var(d$grp2)/length(d$grp2)))
t_value

2*pt(t_value, df = 38)

t.test(d$grp1, d$grp2, var.equal = T, alternative = "less")
pt(t_value, df = 38)
```

## Exercise: t-test

- simulate 2 datasets, n = 50, means = 2 and 5, SDs = 5
- do t-test by hand
- write a result sentence

```{r}
d <- tibble(grp1 = rnorm(50, m = 2, sd = 5),
            grp2 = rnorm(50, m = 5, sd = 5))

t_value <- (mean(d$grp1) - mean(d$grp2))/(sqrt(var(d$grp1)/length(d$grp1) + var(d$grp2)/length(d$grp2)))
t_value

2*pt(t_value, df = 98)
```

Given a critical value of 0.05, the difference in memory performance between two subject groups is not statistically significant (t(98) = -0.513, p = 0.609).

# Date: 02/09/2022

```{r}
p.values <- rep(NA, length.out = 1000)

for(i in 1:1000) {
  sample1 = rnorm(20, mean = 2, sd = 3)
  sample2 = rnorm(20, mean = 5, sd = 3)
  
  p <- t.test(sample1, sample2)$p.value
  p.values[i] = p
}

ggplot() +
  geom_histogram(aes(x = p.values), color = "black", binwidth = .01) +
  coord_cartesian(xlim = c(0, .5))
```

## p-value distribution of null hypothesis

```{r}
p.values <- rep(NA, length.out = 1000)

for(i in 1:1000) {
  sample1 = rnorm(20, mean = 0, sd = 3)
  sample2 = rnorm(20, mean = 0, sd = 3)
  
  p <- t.test(sample1, sample2)$p.value
  p.values[i] = p
}

ggplot() +
  geom_histogram(aes(x = p.values),
                 color = "black",
                 fill = "lightgreen",
                 binwidth = .01)
```

## statistical power
Probability of the detecting the effect when it exists

```{r}
p.values <- rep(NA, length.out = 1000)

for(i in 1:1000) {
  sample1 = rnorm(45, mean = 20, sd = 5)
  sample2 = rnorm(45, mean = 23, sd = 5)
  
  p <- t.test(sample1, sample2)$p.value
  p.values[i] = p
}

mean(p.values < .05)
```

## Exercise: power with varying sample size

plot power across sample sizes (10 to 100)

```{r}
sample.sizes <- seq(from = 10, to = 100, by = 5)
powers <- rep(NA, length.out = 19)

for (i in 1:19) {
  
  p.values <- rep(NA, length.out = 1000)

  for (j in 1:1000) {
    sample1 = rnorm(sample.sizes[i], mean = 20, sd = 5)
    sample2 = rnorm(sample.sizes[i], mean = 23, sd = 5)
    
    p <- t.test(sample1, sample2)$p.value
    p.values[j] = p
  }
  
  powers[i] <- mean(p.values < .05)
}

power.df <- data.frame(sample.size = sample.sizes, power = powers)

power.df %>% 
  ggplot(aes(x = sample.size, y = power)) +
  geom_point() +
  geom_line()
```

